---
title: "Adri's Final Project"
subtitle: "ECO602 Fall 2022"
author: "Adri Grow"
date: "12/16/2022"
output: html_document
---
[Final Project: Index](index.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Reference Guide {.tabset .tabset-pills}


## Loading Data and Packages
To load the palmer penguins dataset, you must load it into your R workspace. 

* `library()` and `require()` are both functions which ensure the package is installed and can be used

```{r, message = FALSE}
# Load and require the here package
library(here)
require(here)
# Load and require the palmerpenguins package
library(palmerpenguins)
require(palmerpenguins)
# Load the penguin data from the palmerpenguins package into a data frame called penguins
penguins <- palmerpenguins::penguins
```

To load a dataset into R, we'll use the `here()` and `read.csv()` functions. 

* `here()` tells R where we want to access the files from on the physical computer

* `read.csv()` transforms the raw data file into an R `data.frame` of columns and rows

* we'll use the `<-` assignment operator to name our data to a `data.frame` called `ginkgo`
```{r}
ginkgo <- read.csv(here("./docs/ginkgo_data_2022.csv"))
```

## Data structures
The function `c()` combines or concatenates its arguments into a vector (a 1-dimensional data structure consisting of 1 or more elements).

All of the elements must be of the same type.
I can’t combine character and numeric types in the same call to `c()`
Here’s two examples using numeric and character data types:

```{r}
## Create a vector of numbers:
num_vec  = c(1, 4, 8, 9, 13)

## Create a vector of characters:
char_vec = c("a", "fish", "data is cool")
```

I can show the contents of a vector by typing the name of the vector, or using the `print()` function.

```{r}
## Typing the name of the vector into the console prints the contents
num_vec
```

```{r}
## The print() function accomplishes the same task:
print(char_vec)
```

The function `length()` can report of set the length of a vector (or a factor or other R objects). I can find the length of the vectors we made above with the `length()` function:
```{r}
length(num_vec)
length(char_vec)
```

The function `matrix()` creates a matrix of values provided. I've made a matrix with the numbers 1-10 in a single column saved in the `data.frame` called `matrix1`
```{r}
matrix1 <- matrix(1:10, nrow = 10)
print(matrix1)
```

The function `data.frame` creates a data frame of variables organized in columns and rows. This is the most basic R structure used.

*Inside the `data.frame()` function, I'm assigning values (contents of rows) to column names
```{r}
protists <- data.frame(
  Cell = c(1:5),
  Organism = c("Foram", "Ciliate", "Testate Amoeba", "Diatom", "Slime mold"),
  Typically_shelled = c("yes", "no", "yes", "yes", "no"),
  Cute = c("yes", "yes", "yes", "yes", "yes")
)
print(protists)
```

Using the Ginkgo 2022 that we loaded above, we can use the `nrow()`, `ncol()`, and `dim()` functions to get information about the data.

* `nrow()` gives the number of rows in the data frame

* `ncol()` gives the number of columns in the data frame

* `dim()` gives the dimensions of the data frame object
```{r}
nrow(ginkgo)
ncol(ginkgo)
dim(ginkgo)
```

## Subsetting
I will use the `ginkgo` data frame to show an example of how we can extract a specific column of the data (called subsetting). I'm saving the subsetted data into a new data frame called `ginkgo_seeds`
```{r}
ginkgo_seeds <- data.frame(ginkgo$seeds_present)
```
I can also subset the data frame to target a specfic range of data. In the next three examples, I will select the first row, the element in row 2, column 3, and then the third column

```{r}
ginkgo_1st_row <- data.frame(ginkgo[1,])
print(ginkgo_1st_row)

ginkgo_row2col3 <- data.frame(ginkgo[2,3])
print(ginkgo_row2col3)

ginkgo_3rd_col <- data.frame(ginkgo[3])
head(ginkgo_3rd_col)
```
Note: I used the `head()` function to view the first few rows of the `ginkgo_3rd_col` data frame (rather than printing the entire column with 220 rows)

Next, I'll use the `subset()` function to select all the Adelie penguin data from the penguins data frame we created earlier with the palmerpenguins package.

* `subset()` returns the data that meets the specified conditions

I'm using `==` to specify that I only want the data for species that are Adelie penguins.

```{r}
Adelie_penguins <- subset(penguins, species=="Adelie")
head(Adelie_penguins)
```

## Numerical Data Exploration
Next I'll use the `summary()`, `mean()`, and `sd()` functions to get some descriptive statistics for the ginkgo dataset. I'm specifically interested in the maximum width of the ginkgo leaves variable.

* `summary()` provides a summary of the data specified, for numeric variables this includes the minimum, median, mean, maximum

* `mean()` calculates the mean for the data specified

* `sd()` calculates the standard deviation for the data specified

```{r}
summary(ginkgo$max_width)
mean(ginkgo$max_width)
sd(ginkgo$max_width)
```

## Graphical Data Exploration
I will create a scatterplot of the ginkgo data. I'm interested in looking at the relationship between maximum leaf depth and maximum leaf width

* the `plot()` function is used to make a scatterplot

* `col=` change the color of the points (color name must be in quotes)

* `pch=` change the shape of the points

* `cex=` change the size of the points

* `main=` add a title to the plot

* `xlab=` add an x-axis label

* `ylab=` add a y-axis label

* `xlim=` set the x-axis coordinates

* `ylim=` set the y-axis coordinates

```{r}
plot(ginkgo$max_depth, ginkgo$max_width, 
     col="darkgreen", 
     pch = 8, 
     cex = 1, 
     main = "Ginkgo leaves: maximum depth x maximum width", 
     xlab = "Maximum Leaf Depth (mm)", 
     ylab = "Maximum leaf width (mm)", 
     xlim=c(0,125), 
     ylim=c(0,150))
```

I will create a histogram next with the palmer penguin data, specifically looking at the flipper lengths.

* the `hist()` function is used to make a historgram

* the `breaks()` functions is used to specify the number of bins you want the data separated into

```{r}
hist(penguins$flipper_length_mm, breaks = 30, xlab="Flipper length (mm)", main="Histogram of penguin flipper length")
```

I will create a simple boxplot next with the ginkgo data, to specifically visualize the ginkgo petiole length.

* the `boxplot()` function is used to make a boxplot

```{r}
boxplot(ginkgo$petiole_length, ylab = "Petiole length (mm)", main = "Ginkgo simple boxplot")
```

Next, I'll make a conditional boxplot so that we can visualize a numerical variable (`max_width`) against a nominal variable (`seeds_present`).

```{r}
boxplot(max_width ~ seeds_present, data = ginkgo, main = "Ginkgo conditional boxplot", xlab = "Seeds", ylab = "Maximum width (mm)")
```

Finally, I'll make a panel of histograms so that we can visualize many plots of the ginkgo data at once.

* the `par()` function is used to set graphical parameters. In our case we want a 2x2 grid and will use the `mfrow()` function argument to specify 2 columns and 2 rows

```{r}
par(mfrow=c(2,2))
hist(ginkgo$max_width, xlab = "Maximum width (mm)", main = "Ginkgo leaf width")
hist(ginkgo$max_depth, xlab = "Maximum depth (mm)", main = "Ginkgo leaf depth")
hist(ginkgo$notch_depth, xlab = "Notch depth (mm)", main = "Ginkgo leaf notch")
hist(ginkgo$petiole_length, xlab = "Petiole length (mm)", main = "Ginkgo leaf petiole")
```

## Distribution Functions

I'll next show examples for six different distribution functions.

The three normal distribution functions are:

* `dnorm()` provides the density; dnorm(x, mean, sd) 

  * use this if you want to know if a specific value is more likely

* `pnorm()` provides the distribution function; pnorm(q, mean, sd)

  * use this if you want to know the probability of a value within a specified range

* `qnorm()` provides the quantile function; qnorm(p, mean, sd)

  * use this if you want to ask about a certain percentile

```{r}
# What is the value of the standard normal distribution at x=1 with a normally-distributed population with mean = 3 and standard deviation = 1.7
dnorm(1, 3, 1.7)

# What is the probability of observing a value of less than 1.5 from a normally-distributed population with mean = 3 and standard deviation = 1.7?
pnorm(1.5, 3, 1.7)

# find the 95th percentile of the standard normal distribution
qnorm(.95, 3, 1.7)
```

The three binomial distribution functions are:

* `dbinom()` provides the probability mass function; dbinom(x, size, prob) 

  * use this if you want to know the probability of getting a value of exactly `x`

* `pbinom()` provides the cumulative mass function; pbinom(q, size, prob) 

  * use this if you want to know the probability of getting a value of `x` or less or more

* `qbinom()` provides the quantile function; qbinom(q, size, prob)

  * use this if you want to ask about a certain percentile

```{r}
# Adri makes 70% of their free-throw attempts. If they shoot 15 free throws, what is the probability that they makes exactly 10?
# we want to find the probability of 10 successes during 15 trials where the probability of success on each trial is 0.7
dbinom(10, 15, 0.7)

# Jailene flips a coin 10 times. What is the probability that the coin lands on heads more than 5 times?
# we want to find the probability of more than 5 successes during 10 trials where the probability of success on each trial is 0.5
pbinom(5, size=10, prob=.5, lower.tail=FALSE)

# we want to find the 90th percentile of a binomial distribution with 50 trials and probability of success for each trial is 0.50
qbinom(.90, size=50, prob=.50)
```

