---
title: "Adri's Final Project"
subtitle: "ECO602 Fall 2022"
author: "Adri Grow"
date: "12/21/2022"
output: html_document
---
[Final Project: Index](index.html)

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# R Reference Guide {.tabset .tabset-pills}


## Sample Code Example

The function `c()` combines or concatenates its arguments into a vector (a 1-dimensional data structure consisting of 1 or more elements).

All of the elements must be of the same type.
I can’t combine character and numeric types in the same call to `c()`
Here’s two examples using numeric and character data types:

```{r}
## Create a vector of numbers:
num_vec  = c(1, 4, 8, 9, 13)

## Create a vector of characters:
char_vec = c("a", "fish", "data is cool")
```

I can show the contents of a vector by typing the name of the vector, or using the `print()` function.

```{r}
## Typing the name of the vector into the console prints the contents
num_vec
```

```{r}
## The print() function accomplishes the same task:
print(char_vec)
```

## Loading Data and Packages


## Data structures

## Subsetting

## Numerical Data Exploration

## Graphical Data Exploration

## Distribution Functions







